### 2025-04-21 sec/3_tls_opt.tex
Shorten ineq solver

We can show that this optimization problem is equivalent to the following problem that finds the optimal axis of rotation $\mathbf{n}^*$:

@@

with the data matrices $\mathbf{A}$ and $\mathbf{g}$ that depend on the points $\mathbf{a}_i$ and $\mathbf{b}_i$ and the angle constraint $n_r$:

\begin{equation}
	\label{eq:davenport-data-matrices}
	\begin{aligned}
		\mathbf{B} &= \sum_{i=1}^N w_i \mathbf{R}_c^\transposed \mathbf{a}_i\mathbf{b}_i^\transposed \in \Rone^{3 \times 3}, \quad
		\mathbf{z} = \sum_{i=1}^N w_i [\mathbf{a}_i]_{\times}\mathbf{b}_i \in \Rone^{3}, \quad \mathbf{C} = \mathbf{B} + \mathbf{B}^\transposed\\
		\mathbf{A} &= \sin^2 \left(\frac{n_r}{2} \right) \mathbf{C}, \quad \mathbf{g} =  \sin \left(\frac{n_r}{2} \right) \cos \left(\frac{n_r}{2} \right) \mathbf{z}
	\end{aligned}
\end{equation}

This derivation is based on the Daveport's Quaternion-method \cite{davenportsQ} \cite{8594296}, see \ref{proof:wls-relax-eq-constrained-rotation} for a proof of the equivalence of both optimization problems.

To solve the optimization problem \ref{eq:eq-constrained-prob-ev}, we perform overall two steps: We first need to compute the eigenvalue decomposition  $\mathbf{A} = \mathbf{Q}\mathbf{\Lambda}\mathbf{Q}^\transposed$ and then need to find largest root of the following rational function $C(\gamma)$ (characteristic polynomial):

\begin{equation}
	\label{eq:characterisitc-poly}
	\begin{aligned}
		C(\gamma) = \sum_{i=1}^3 \left(\frac{a_i}{\lambda_i - \gamma}\right)^2 - 1
	\end{aligned}
\end{equation}

with
\begin{equation}
	\begin{aligned}
		\mathbf{a} = (a_1, a_2, a_3)^\transposed = \mathbf{Q}^\transposed \mathbf{g} \in \Rthree, \quad \mathbf{\Lambda} = \text{diag}(\lambda_1, ..., \lambda_3)
	\end{aligned}
\end{equation}

\begin{algorithm}[!ht]
	\caption{Solve constrained eigenvalue problem}
	\label{alg:solve-constrained-ev}
	\begin{algorithmic}[1]
		\Require{Data matrices $\mathbf{A} \in \Rone^{3\times3}, \mathbf{g} \in \Rthree$}
		\Ensure Solution $\mathbf{n}^*$ to the optimization problem Eq. \ref{eq:eq-constrained-prob-ev}
		\hypertarget{ref:min-constrained-ev}{
			\Function{\code{solve\_constrained\_ev}}{$\mathbf{A}, \mathbf{g}$}}
		\State $\code{kEps} \leftarrow 2^{-48}$ \algorithmiccomment{Epsilon-value for double precision}
		\State $\mathbf{Q}, \mathbf{\Lambda} \leftarrow \code{eig(} \mathbf{A} \code{)}$ \algorithmiccomment{Compute eigen-decomposition}
		\State $k = \argmax\{\lambda_1, \lambda_2, \lambda_3\}$
		\State $\mathbf{a} \leftarrow \mathbf{Q}^\transposed \mathbf{g}$
		\State $l = \lambda_k + |a_k| - \code{kEps}, r = ||\mathbf{a}|| + \lambda_k + \code{kEps}$ \algorithmiccomment{Bracket root}
		\State $\hat{\gamma} \leftarrow \code{bisect} (C, [l, r] \code{)}$ \algorithmiccomment{Find root of Eq.\ref{eq:characterisitc-poly} with bisection}
		\State $\mathbf{f} \leftarrow \left(\frac{a_1}{\lambda_1 - \hat{\gamma}}, \frac{a_2}{\lambda_2 - \hat{\gamma}}, \frac{a_3}{\lambda_3 - \hat{\gamma}} \right)$
		\State $\mathbf{n}^* \leftarrow \mathbf{Q}\mathbf{f}$ 
		\State \Return $\mathbf{n}^*$
		\EndFunction
	\end{algorithmic}
\end{algorithm}

This algorithm is shown in Alg \ref{alg:solve-constrained-ev}, see \ref{sec:constrained-ev-problem-derivation} for a derivation with explanations.
The eigen-decomposition can exploit the fact that the matrix is real and symmetric (but not positive semi-definite) for which faster algorithms are available (i.e. Eigen's \code{Eigen::SelfAdjointEigenSolver}). Note also that the vector $\mathbf{f}$ in line 8 is equal to $(\mathbf{\Lambda} - \hat{\gamma} \mathbf{I})^{-1} \mathbf{a}$. No actual matrix-inversion is required however since the matrix is diagonal and also with floating-point arithmetic it is numerically more stable/accurate to compute $a/b$ instead of $\frac{1}{b} a$.

\subsection{Overall active-set solver}

\begin{algorithm}[!ht]
	\caption{Active-set method for minimizing the WLS-relaxation over rotations (\ref{eq:tls-rwls})}
	\label{alg:min-wls-relax-rot}
	\begin{algorithmic}[1]
		\Require{Point cloud matrices $\mathbf{A}, \mathbf{B}$, node interval $(\mathbf{R_c}, n_r)$, computed WLS-relaxation weights vector $\mathbf{w}$}
		\Ensure Solution to the optimization problem \ref{eq:tls-rwls}
		\hypertarget{ref:min-wls-relax-rot2}{
			\Function{\code{Minimize\_WLS\_Relax\_Rotation}}{$\mathbf{A}, \mathbf{B}, \mathbf{R_c}, n_r, \mathbf{w}$}}
		\State $\mathbf{B} \leftarrow \sum_{i=1}^N w_i \mathbf{R}_c^\transposed \mathbf{a}_i\mathbf{b}_i^\transposed $ \algorithmiccomment{Compute cross-correlation matrix}
		\State $\bar{\Delta \mathbf{R}} \leftarrow \code{svd\_rot(} \mathbf{B} \code{)}$  \algorithmiccomment{Solve unconstrained with SVD-algorithm}
		\If{$d_{\angle}(\mathbf{I}, \Delta\mathbf{R}) \leq n_r$} \algorithmiccomment{Check if constraint is already satisfied}
		\State \Return $\mathbf{R_c} \bar{\Delta \mathbf{R}}$
		\EndIf
		\State $\mathbf{A}, \mathbf{g} \leftarrow \code{get\_davenport\_matrices(} \mathbf{B}, n_r \code{)}$ \algorithmiccomment{Get  data-matrices (Eq. \ref{eq:davenport-data-matrices})}
		\State $\mathbf{n}^* \leftarrow \code{solve\_constrained\_ev(} \mathbf{A}, \mathbf{g} \code{)}$ \algorithmiccomment{Solve for
			rotation axis (Eq. \ref{eq:eq-constrained-prob-ev})}
		\State $\bar{\Delta \mathbf{R}} \leftarrow \code{from\_axis\_angle(}\mathbf{n}^*, n_r \code{)}$ \algorithmiccomment{ Construct optimal rotation}
		\State \Return $\mathbf{R_c} \bar{\Delta \mathbf{R}}$
		\EndFunction
	\end{algorithmic}
\end{algorithm}

Our active-set solver for finding the global minimum of the WLS-relaxation over rotations is shown in Alg. \ref{alg:min-wls-relax-rot}. It is very efficient since the only computationally intensive operation is the computation of the $3 \times 3$ cross-correlation matrix $\mathbf{B}$ (line 13) that has linear runtime in the number of points. All the operations that follow are based on this $3 \times 3$ matrix and therefore have a constant runtime (i.e. independent of the number of points) and are in practice very fast. For example, the eigen-decomposition (line 3) takes only around $10 \mu s$ while the computation of the cross-correlation matrix requires aroundd $500 \mu s$ for 10k points. This is also because the cross-correlation matrix is re-used between the SVD-algorithm (line 3) and the algorithm for the equality-constraint case (line 7). The function \code{from\_axis\_angle} converts a given axis and angle to a rotation matrix using Rodrigues' formula.

This solver is novel to our knowledge -- the problem of solving the Wahba-problem with inequality-constrained angle has not been proposed in the literature before.

The solver for the equality-constraint case ($\code{solve\_constrained\_ev}$) is implemented from the derivation first done in \cite{10.1007/978-3-642-75536-1_57} and inspired by the \textit{Trust-Region Subproblem} (TRS) that arises in trust-region (local) optimization algorithms \cite{Adachi2017SolvingTT} such as Levenberg-Marquardt. For such TRS-problems, an active-set method has been proposed also in \cite{Rontsis2022} where it was generalized additionally for a minimum-norm constraints.

>>>>

We show that this optimization problem is equivalent to the following problem that finds the optimal axis of rotation $\mathbf{n}^*$:

@@

with the data matrices $\mathbf{A}$ and $\mathbf{g}$ that depend on the points $\mathbf{a}_i$ and $\mathbf{b}_i$ and the angle constraint $n_r$. A definition is given in the Appendix \ref{thm:constrained-wahba-davenport}.

To solve this non-convex optimization problem \ref{eq:eq-constrained-prob-ev} over the sphere, we use the algorithm proposed in \cite{10.1007/978-3-642-75536-1_57}. We perform overall two steps: We first need to compute the eigenvalue decomposition  $\mathbf{A} = \mathbf{Q}\mathbf{\Lambda}\mathbf{Q}^\transposed$ and then need to find the rightmost root of a certain rational function (characteristic polynomial). This can be done in practice very fast, yielding a constant time algorithm.

Our active-set solver for finding the global minimum of the WLS-relaxation over rotations is is overall very efficient. The only computationally intensive operation is the computation of the $3 \times 3$ cross-correlation matrix $\mathbf{B}$ (line 13) that has linear runtime in the number of points. All the operations that follow are based on this $3 \times 3$ matrix and therefore have a constant runtime (i.e. independent of the number of points). (For example, the eigen-decomposition (line 3) takes only around $10 \mu s$ while the computation of the cross-correlation matrix requires aroundd $500 \mu s$ for 10k points.)
This solver is novel to our knowledge -- the problem of solving the Wahba-problem with inequality-constrained angle has not been proposed in the literature before.

### 2025-04-21 sec/4_evaluation.tex
Shorten ineq solver

\section{Experiments}
In this section, we present experiments where we study the intersection graphs arising from the robust point cloud registration problem. We evaluate the runtime of state-of-the-art maximum clique finding algorithms.

\label{sec:evaluation}
We perform the experiments on a desktop PC with an Intel i7 12700KF CPU and 32GiB of RAM.
\subsection{Synthetic experiments}

@@

We generate synthetic point clouds by sampling inliers from a unit cube $s \cdot [-1, 1]^3$ with scale $s$. We add uniform measurement noise by sampling a ball of radius $\epsilon^2$, and then sample outliers in the bounding box of the inliers, also uniformly. First, we set $s=10$ and $\epsilon=0.5$. As shown in Fig. \ref{fig:execution-time-over-outlier-rate}, the runtime is very low for $N =100$ and decreases until 90\% outlier-rate because the graph sparsity increases. Above 90\% outlier-rate, it increases because the number of visited cliques increases.

@@

\subsubsection{Validity of the global optimality certificate}
Even though GNC is a heuristic, based on our case analysis (Sec. \ref{sec:branching-over-comb}) it seems likely that GNC finds global minimum under certain conditions.
We therefore test on synthetic data whether the global optimality certificate is valid, i.e. whether at termination, the lower bound is less or equal than the TLS objective at the ground-truth solution.
We define the suboptimality $\eta$ of the solution based on the upper bound (UB), that is, the best known solution, and the lower bound (LB) \cite{9785843}
\begin{equation}
	\begin{aligned}    
		\eta = \frac{\text{UB} - \text{LB}}{1 + |\text{UB}| + |\text{LB}|}
	\end{aligned}
\end{equation}
We test with $N=100$, $\epsilon=0.5$ and vary the data scale $s$ from 3 to 100. We test for each scale outlier rates of $0\%, 30\%, 50\%, 90\%, 96\%$, and do 1000 trials for each combination.
\begin{figure}[!ht]
	\centering
	\includegraphics[width=0.49\linewidth]{figures/certificate_validity_vs_scale}
	\includegraphics[width=0.49\linewidth]{figures/eta_subopt_over_outlier_rate}
	\caption{Amount of instances with valid global optimality certificate (i.e. valid suboptimality) on synthetic data, including achieved $\eta$-suboptimality.}
	\label{fig:cert-validity}
\end{figure}
The results are shown in Fig. \ref{fig:cert-validity}, we clearly see that with a larger data scale, the likelihood of a valid certificate (and successful registration result) increases. While this observation alone could be explained solely by the solver being able to solve the problem more likely, we also see that the suboptimality increases for instances above 90\% outliers. This indicates that our solver is able to detect that some instances could not be solved to global optimality.
Unfortunately, being only able to reject some and not all instances means our solver is currently neither globally optimal (being able to find the global minimum of every instance and outputting a valid lower bound on every instance), nor certifiable (outputting only a valid lower bound on every instance).

>>>>

\section{Experimental evaluation}
\label{sec:evaluation}
In this section we evaluate our Branch and Bound solver that uses our novel convex relaxation on point cloud registration and related outlier robust problems. 
We measure the required number of iterations as well as the per-iteration costs. We have implemented a simple version of a best-first Branch-and-Bound \cite{10.1007/978-3-642-37444-9_42}.

We also have validated the corecteness of the algorithm and it's implementation empirically: We did not find any instance in which it did not find the global minimum of the TLS problem.

We perform the experiments on a desktop PC with an Intel i7 12700KF CPU and 32GiB of RAM.

@@

\begin{figure}[!ht]
	\centering
	\begin{adjustbox}{width=1.\linewidth}
		\includegraphics{figures/robust_wahba_runtime_rotation_error.pdf}
	\end{adjustbox}
	\caption{Der durchschnittliche Rotationsfehler in Abhängigkeit vom Anteil der Ausreißer: Unsere Methode bleibt robust bis zu 95\% Ausreißer, wie von der Truncated Least Squares Formulierung zu erwarten ist.}
	\label{fig:rotation-error-over-outlier-rate}
\end{figure}

\begin{figure}[!ht]
	\centering
	\begin{adjustbox}{width=1.\linewidth}
		\includegraphics{figures/robust_wahba_runtime_execution_time.pdf}
	\end{adjustbox}
	\caption{Die durchschnittliche Ausführungszeit von Branch-and-Bound in Abhängigkeit von der Gesamtzahl der Punkte $N$ des Optimierungsproblems.}
	\label{fig:runtime}
\end{figure}

We generate synthetic point clouds by sampling inliers from a unit cube $s \cdot [-1, 1]^3$ with scale $s$. We add uniform measurement noise by sampling a ball of radius $\epsilon^2$, and then sample outliers in the bounding box of the inliers, also uniformly. First, we set $s=10$ and $\epsilon=0.5$. As shown in Fig. \ref{fig:execution-time-over-outlier-rate}, the runtime is very low for $N =100$.

The results of the rotation error as a function of the outlier rate are shown in Fig. \ref{fig:rotation-error-over-outlier-rate} is shown. The estimation remains correct with a small error of less than $1$ degree. We also investigated the scalability of our method over the total number of points $N$. As the results in Fig. \ref{fig:runtime} show, our approach can even be used for real-time applications with a runtime of less than 100 milliseconds.

@@

\subsubsection{Validity of the global optimality certificate}

We test on synthetic data whether the global optimality certificate is valid, i.e. whether at termination, the lower bound is less or equal than the TLS objective at the ground-truth solution.
We define the suboptimality $\eta$ of the solution based on the upper bound (UB), that is, the best known solution, and the lower bound (LB) \cite{9785843}
\begin{equation}
	\begin{aligned}    
		\eta = \frac{\text{UB} - \text{LB}}{1 + |\text{UB}| + |\text{LB}|}
	\end{aligned}
\end{equation}
We test with $N=100$, $\epsilon=0.5$ and vary the data scale $s$ from 3 to 100. We test for each scale outlier rates of $0\%, 30\%, 50\%, 90\%, 96\%$, and do 1000 trials for each combination.
\begin{figure}[!ht]
	\centering
	\includegraphics[width=0.49\linewidth]{figures/certificate_validity_vs_scale}
	\includegraphics[width=0.49\linewidth]{figures/eta_subopt_over_outlier_rate}
	\caption{Amount of instances with valid global optimality certificate (i.e. valid suboptimality) on synthetic data, including achieved $\eta$-suboptimality.}
	\label{fig:cert-validity}
\end{figure}

TODO
